Estimating duration in the presence of misclassification

### 2024-02-14 transmission.tex
resolve all Overleaf comments

\todo[inline]{Intro probably needs rewriting, with novelty claims and a bit more context. Revisit once intro chapters written.}

In this chapter, I use the preferred estimate of the duration of RT-PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to investigate estimation of incidence and transmission of SARS-CoV-2 in England, solely on the basis of .
CIS data (as shown in \cref{E-biology-data:sec:positivity} for further details).
I will show that the CIS data is sufficient to derive such estimates, suggesting surveys of this type would be sufficient for surveillance of an epidemic.

I take two approaches, the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
First, in \cref{backcalc}, I use a back-calculation approach, based on \cref{E-inc-prev:eq:EPt}. This approach enables estimation of incidence within (sub)populations of England. Then, in \cref{SEIR}, I use a mechanistic compartmental model, a variation of the model introduced in \cref{E-SEIR:sec:mechanistic-models}, to estimate both incidence and components of transmission, \eg and whether changes in transmission can be explained by changes in contact rates.
Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.

\todo[inline]{
    Lit review to go in introduction chapter, framing the whole thesis because it is important in understanding the motivation for the whole work.
    Would include \textcite{mccabeCISincidence,abbottCISincidence} on backcalculation. Also Taffe?
    As well as the literature on mechanistic models using REACT and CIS I have previously circulated.
}
\section{Data} \label{transmission:sec:data}

@@

During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
There is weak evidence that when incidence starts increasing is correlated with when the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for all of England (A) and its component regions.
        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
        Yorkshire and the Humber is labelled as Yorkshire for brevity.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}
\begin{figure}
    \includegraphics{transmission/backcalc-alpha}
    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
        Linear regression line shown.
    }
\end{figure}

The incidence proportion was highest among the 11--15 and 16--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information for the posterior to overcome the prior penalisation towards this situation.
The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
There are many plausible explanations for this effect.
The higher incidence in this age group prior to November means it could be an immunity effect.
Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
Many other explanations are possible, and further work is required to determine causality.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
    \caption[Incidence estimated using backcalculation by age group]{%
        Incidence proportion for each age group, on both natural and log scales.
    }
    \label{transmission:fig:backcalc-ages}
\end{figure}

@@

In this section, I apply the general mechanistic transmission modelling framework from \cref{E-SEIR} to the CIS prevalence data.
The application of this framework follows \textcite{birrellRealtime}, with the most substantial context-specific work in parameterising the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other studies collecting representative prevalence data, \eg REACT.
In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
Having specified the full model (\cref{SEIR:sec:methods-application}), I conducted a simulation study to validate the inference process and identifiability (\cref{SEIR:sec:sim-study}), before applying the model to the CIS data (\cref{SEIR:sec:application}).
In \cref{SEIR:sec:discussion}, I discuss the results; in \cref{transmission:sec:comparison} I compare the results to those in the previous section.

@@

This section  explains the novel observation model approach I propose for fitting to the CIS prevalence data.
This will link the model in the previous section to the CIS prevalence data.

% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model (see \cref{E-SEIR:sec:SIR}), and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.

>>>>

\todo[inline]{%
    Intro probably needs rewriting, with novelty claims and a bit more context.
    Revisit once intro chapters written.
    Question: is CIS (or similar) enough to estimate incidence / transmission?

}

In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate estimation of incidence and transmission of SARS-CoV-2 in England.
This estimation is based only on the CIS data (as shown in \cref{E-biology-data:sec:positivity} for further details).
I will show that the CIS data is sufficient; demonstrating the value of surveys of this type for pandemic surveillance.

I take two approaches, backcalculation and a mechanistic model, the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
\todo{what are the advantages of each approach specifically in this context?}

First, in \cref{backcalc}, I use a back-calculation approach, based on \cref{E-inc-prev:eq:EPt}.
This approach enables estimation of incidence within (sub)populations of England.

In \cref{SEIR}, I use a mechanistic compartmental model, within the modelling framework introduced in \cref{E-SEIR:sec:mechanistic-models}.
This approach enables estimating both incidence and components of transmission.
Estimating transmission in addition to incidence answers important epidemiological questions, \eg whether changes in transmission can be explained by changes in contact rates.
Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.

\section{Data} \label{transmission:sec:data}

@@

During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs.
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for all of England (A) and its component regions.
        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
        Yorkshire and the Humber is labelled as Yorkshire for brevity.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}

The incidence proportion was highest among the 11--15 and 16--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information in the data to overcome the prior which favoured this situation.
The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
There are many plausible explanations for this effect.
The higher incidence in this age group prior to November means it could be an immunity effect.
Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
Many other explanations are possible, and further work is required to determine causality.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
    \caption[Incidence estimated using backcalculation by age group]{%
        Incidence proportion for each age group, on both natural and log scales.
    }
    \label{transmission:fig:backcalc-ages}
\end{figure}

@@

In this section, I apply the general mechanistic transmission modelling framework from \cref{E-SEIR} to the CIS prevalence data.
The application of this framework follows \textcite{birrellRealtime}, with the most substantial context-specific work\todo{maybe change ``work" to something else: I want a succinct way to say that this was the major part that needed changing from the generic framework in chapter 3} in parameterising the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other studies collecting representative prevalence data, \eg REACT.
In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
Having specified the full model (\cref{SEIR:sec:methods-application}), I conducted a simulation study to validate the inference process and identifiability (\cref{SEIR:sec:sim-study}), before applying the model to the CIS data (\cref{SEIR:sec:application}).
In \cref{SEIR:sec:discussion}, I discuss the results; in \cref{transmission:sec:comparison} I compare the results to those in the previous section.

@@

\todo[inline]{Rewrite this intro paragraph}
This section  explains the novel observation model approach I propose for fitting to the CIS prevalence data.
This will link the model in the previous section to the CIS prevalence data.

% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model, and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.
